Tree growth, mortality, and climatic stress in west-coast states, USA

### 2025-09-09 Manuscript.Rmd
Created a third table for the manuscript summarizing the two difference figures findings.

# Significant findings, Growth
grow.n <- nrow(grow.diff$AvB2)

@@

# Significant findings, Mort
mort.AvB.sig <- length(mort.diff$AvB2$significant[mort.diff$AvB2$significant == TRUE])
mort.AvB.signeg <- length(mort.diff$AvB2$significant[mort.diff$AvB2$significant == TRUE & mort.diff$AvB2$Means < 0])
mort.A.sig <- length(mort.diff$A_LMH2$significant[mort.diff$A_LMH2$significant == TRUE])
mort.A.signeg <- length(mort.diff$A_LMH2$significant[mort.diff$A_LMH2$significant == TRUE & mort.diff$A_LMH2$Means < 0])
mort.B.sig <- length(mort.diff$B_LMH2$significant[mort.diff$B_LMH2$significant == TRUE])
mort.B.signeg <- length(mort.diff$B_LMH2$significant[mort.diff$B_LMH2$significant == TRUE & mort.diff$B_LMH2$Means < 0])

@@

@@

# 
fig_nums <- captioner::captioner(prefix = "Figure")
tab_nums <- captioner::captioner(prefix = "Table")

f.ref <- function(x) {
 stringr::str_extract(fig_nums(x), "[^:]*")
}

@@

tab.tbl_growth <- tab_nums(name = "tbl_growth", 
                        caption = "Decadal growth estimates by state and overall for analysis species. Mean estimates with confidence intervals and plot number are provided. Estimates are not made for states with fewer than 10 plots.")

@@

fig.domains <- fig_nums(name = "fig_domains",
                        caption = "Decadal (A) growth and (B) mortality domain estimates for *Abies lasiocarpa*.  The six domains are defined as FIA plots that fall [A]bove or [B]elow the CWD change threshold with [L]ow, [M]edium, or [H]igh initial CWD values (see text for details). E.g., 'AL' is Above CWD change threshold, Low initial CWD value. The distribution of plot CWD change and initial values (C) is presented as well, with the blue horizontal line at a change in CWD of 5 mm representing the change threshold and the green vertical lines representing the quantile boundaries between Low, Medium, and High initial CWD values." )

@@

@@

We obtained empirical means and 95% confidence intervals via bootstrap. For every species we sampled plots with replacement from within each of the strata and obtained point estimates.  For each domain we calculated a mean and upper and lower 95% CIs with the 2.5% and 97.5% quantiles from 1000 sampling iterations.  This approach allows for nonsymmetrical CIs around a mean and prevents CIs from including negative values.  We estimated differences between pairs of CWD domains by subtracting within-iteration domain point estimates.    
[permutation analysis]

@@

We examined differences in 10-year growth estimates among domains (Figure `r fig_nums("fig_diffgrow", display = "num")`).  The column "Above vs. Below Threshold" compares the domains above the CWD change threshold with those below it (e.g., for *Abies lasiocarpa*, subtracting the estimates for columns BL, BM, and BH from AL, AM, AH, respectively; Figure `r fig_nums("fig_domains", display = "num")`A).  There were `r grow.AvB.sig` estimates out of `r grow.n` which had 95% CIs that did not overlap with zero; `r grow.AvB.signeg` of these were negative. Negative estimates indicate that the growth rate was greater for a species in plots below the change estimates.

The second column of Figure `r fig_nums("fig_diffgrow", display = "num")` shows the differences in growth estimates AH - AL, AH - AM, and AM - AL for each species; effectively, the top row of domains shown in Figure `r fig_nums("fig_domains", display = "num")`A.  `r grow.A.sig` comparisons had 95% CIs that did not include 0, of which `r grow.A.signeg` were negative.  Negative values indicate that a lower-valued initial CWD domain had a greater growth rate than a higher-valued domain.  The third column displays the same domain differences but for the bottom row of domains shown in `r fig_nums("fig_domains", display = "num")`A.  `r grow.B.sig` comparisons had 95% CIs that did not include 0, of which `r grow.B.signeg` were negative. Like the second column, negative differences indicate lower initial-value CWD domains had higher growth than higher domains.

We examined mortality estimates in a similar fashion.  The first column in Figure `r fig_nums("fig_diffmort", display = "num")` has `r mort.AvB.sig` comparisons with 95% CIs that do not overlap zero, `r mort.AvB.sig - mort.AvB.signeg` of which are positive, indicating mortality is greater for those species in one or more of the above-threshold CWD change domains.  The middle column had `r mort.A.sig` "significant" comparisons, `r mort.A.signeg` of which were negative (for the above-threshold change CWD domains, there was more mortality for lower initial-value CWD domains).  The right column had `r mort.B.sig` "significant" comparisons, `r mort.B.signeg` of which were negative (for the below-threshold change CWD domains, there was more mortality for lower initial-value CWD domains).

@@

```{r mapfig, fig.height = 7,  fig.dpi=300, fig.cap=fig.cwdmap_cap, fig.cap.pre="", fig.cap.sep=""}

@@

```{r domainsfig, fig.width = 6,  fig.dpi=300, fig.cap=fig.domains, fig.cap.pre="", fig.cap.sep=""}

@@

```{r diff_grow_fig, fig.height = 6, width = 6, fig.dpi=300, fig.cap=fig.diff_grow, fig.cap.pre="", fig.cap.sep=""}

@@

```{r diff_mort_fig, fig.height = 6, width = 6, fig.dpi=300, fig.cap=fig.diff_mort, fig.cap.pre="", fig.cap.sep=""}

@@

>>>>

# Significant findings, Growth
grow.n.AvB <- nrow(grow.diff$AvB2)
grow.n.A <- nrow(grow.diff$A_LMH2)
grow.n.B <- nrow(grow.diff$B_LMH2)

@@

# Significant findings, Mort
mort.n.AvB <- nrow(mort.diff$AvB2)
mort.n.A <- nrow(mort.diff$A_LMH2)
mort.n.B <- nrow(mort.diff$B_LMH2)

mort.AvB.sig <- length(mort.diff$AvB2$significant[mort.diff$AvB2$significant == TRUE])
mort.AvB.signeg <- length(mort.diff$AvB2$significant[mort.diff$AvB2$significant == TRUE & mort.diff$AvB2$Means < 0])
mort.A.sig <- length(mort.diff$A_LMH2$significant[mort.diff$A_LMH2$significant == TRUE])
mort.A.signeg <- length(mort.diff$A_LMH2$significant[mort.diff$A_LMH2$significant == TRUE & mort.diff$A_LMH2$Means < 0])
mort.B.sig <- length(mort.diff$B_LMH2$significant[mort.diff$B_LMH2$significant == TRUE])
mort.B.signeg <- length(mort.diff$B_LMH2$significant[mort.diff$B_LMH2$significant == TRUE & mort.diff$B_LMH2$Means < 0])

# Summary of significant Growth and Mortality findings for table creation.
datasets.grow <- list(grow.diff$AvB2, grow.diff$A_LMH2, grow.diff$B_LMH2)
datasets.mort <- list(mort.diff$AvB2, mort.diff$A_LMH2, mort.diff$B_LMH2)
datatypes <- list("Above vs. Below", "Above", "Below")
n.comp.grow <- list(grow.n.AvB, grow.n.A, grow.n.B)
n.comp.mort <- list(mort.n.AvB, mort.n.A, mort.n.B)

growth.summ <- pmap(list(datasets.grow, datatypes, n.comp.grow), summ.fig.fcn) %>% bind_rows()
mortality.summ <- pmap(list(datasets.mort, datatypes, n.comp.mort), summ.fig.fcn) %>% bind_rows()

@@

# Adding a section identifier to each table
growth.summ2 <- growth.summ %>%
  mutate(Type = "Growth", .before = 1)

mortality.summ2 <- mortality.summ %>%
  mutate(Type = "Mortality", .before = 1)

# Combine the tables
grow.mort.summ <- bind_rows(growth.summ2, mortality.summ2)

# Create flextable
g.m.summ.ft <- flextable(grow.mort.summ) %>%
  merge_v(j = "Type") %>%  # Merge cells in Section column
  valign(j = "Type", valign = "top") %>%  # Top align section labels
  flextable::align(j = "Type", align = "left") %>%   # Left align section labels
  line_spacing(i = NULL, j = NULL, space = 0.5, part = "body") %>% # Control space between lines. 
  set_table_properties(layout = "autofit", width = 0) %>%
  theme_zebra(
    odd_header = "#FFFFFF",   # White background for all
    even_header = "#FFFFFF",
    odd_body = "#FFFFFF",     
    even_body = "#FFFFFF"     
  ) %>%
    flextable::align(j = 3:7, align = "center", part = "all")

@@

# 
fig_nums <- captioner::captioner(prefix = "")
tab_nums <- captioner::captioner(prefix = "Table")

f.ref <- function(x) {
 caption_text <- fig_nums(x)
  # Remove any leading numbers and extract just "Figure X"
  stringr::str_extract(caption_text, "\\d+")
}

@@

tab.tbl_growth <- tab_nums(name = "tbl_growth", 
                        caption = "Decadal growth estimates by state and overall for analysis species (cm\u00B2/decade). Mean estimates with confidence intervals and plot number are provided. Estimates are not made for states with fewer than 10 plots.")

@@

fig.domains <- fig_nums(name = "fig_domains",
                        caption = "Decadal (A) growth and (B) mortality domain estimates for *Calocedrus decurrens*.  The six domains are defined as FIA plots that fall [A]bove or [B]elow the CWD change threshold with [L]ow, [M]edium, or [H]igh initial CWD values (see text for details). E.g., 'AL' is Above CWD change threshold, Low initial CWD value. The distribution of plot CWD change and initial values (C) is presented as well, with the blue horizontal line at a change in CWD of 5 mm representing the change threshold and the green vertical lines representing the quantile boundaries between Low, Medium, and High initial CWD values." )

@@

tab.summ_grow_mort <- tab_nums(name = "tbl_summ_grow_mort",
                               caption = paste0("Summary of number of comparisons with 95% CIs that do not overlap with zero ([T]otal, [P]ositive-, and [N]egative-valued estimates) for Figures ", fig_nums("fig_diffgrow", display = "num"), " and ", fig_nums("fig_diffmort", display = "num"), " by analysis (T)ype (Growth or Mortality) and type of (C)omparison.  The column Domains lists the domains compared; All is the sum of values within a Comparison. See text for other domain labels.  'n' is the total number of differences examined within a Comparison."))

@@

We obtained empirical means and 95% confidence intervals via bootstrap. For every species we sampled plots with replacement and obtained point estimates.  For each domain we calculated a mean and upper and lower 95% CIs with the 2.5% and 97.5% quantiles from 1000 sampling iterations.  This approach allows for nonsymmetrical CIs around a mean and prevents CIs from including negative values.  We estimated differences between pairs of CWD domains by subtracting within-iteration domain point estimates.    
[permutation analysis]

@@

We examined differences in 10-year growth estimates (Figure `r fig_nums("fig_diffgrow", display = "num")`) and mortality estimates (Figure `r fig_nums("fig_diffmort", display = "num")`) among domains.  The column "Above vs. Below Threshold" compares the domains above the CWD change threshold with those below it (e.g., for *Abies lasiocarpa*, subtracting the estimates for columns BL, BM, and BH from AL, AM, AH, respectively; Figure `r fig_nums("fig_domains", display = "num")`A). Negative estimates for the "Above vs. Below Threshold" in Figures `r fig_nums("fig_diffgrow", display = "num")` and `r fig_nums("fig_diffmort", display = "num")`  indicate that the 10-year growth or mortality rate was greater for a species in plots below the change estimates, respectively.  `r t.ref("tbl_summ_grow_mort")` summarizes the number of domain differences in each of the columns in Figures `r fig_nums("fig_diffgrow", display = "num")` and `r fig_nums("fig_diffmort", display = "num")` where the 95% CI does not include zero as well as the number of comparisons per figure column.

Growth domain comparisons (`r fig_nums("fig_diffgrow")`, `r t.ref("tbl_summ_grow_mort")`) did not appear to be produce substantially more "significant" negative vs. positive estimates, and all three columns of `r fig_nums("fig_diffgrow")` had similar numbers of "significant" comparisons. Growth may have been susceptible (negatively or positively) to a change in CWD and differed by initial CWD values, and the responses are species-specific. *Thuja plicata* exhibited greater growth in the Above and Below Medium domains than in the Lower or Higher domains.  *Abies amablis* had lower growth in the Above and Below High domains than the others, where initial-visit CWD was the greatest.  *Picea englelmannii* had lower growth in the Low domain than the High domain, indicating that more growth was occurring where trees were more water stressed.

Mortality domain comparisons are similar in certain regards to the growth comparisons: results for the three types of domain comparisons did not appear to be predominantly positive or negative (`r fig_nums("fig_diffmort")`, `r t.ref("tbl_summ_grow_mort")`).  There were fewer mortality estimates with "significant" findings than growth. There were fewer differences for "Above vs. Below Threshold" comparisons (`r t.ref("tbl_summ_grow_mort")`) than for the comparisons Above or Below the threshold, indicating that mortality may not be as strongly associated with the change in CWD as by the starting value.  *Abies concolor* had greater mortality at the Above Low and High categories than the Above Middle.  *Calocedrus decurrens* had a higher mortality rate at the High Above domains than the other Above domains. *Picea engelmannii* was the opposite; mortality was lowest at the Above High domain compared to other domains.  *Psuedotsuga menziesii* exhibited greater mortality at its Below High domain than other Below domains.  The two *Quercus* species did not exhibit strong patterns to their mortality.

@@

```{r mapfig, fig.height = 7,  fig.dpi=300, fig.cap=fig.cwdmap_cap, fig.cap.sep=""}

@@

```{r domainsfig, fig.width = 6,  fig.dpi=300, fig.cap=fig.domains, fig.cap.sep=""}

@@

```{r diff_grow_fig, fig.height = 6, width = 6, fig.dpi=300, fig.cap=fig.diff_grow, fig.cap.sep=""}

@@

```{r diff_mort_fig, fig.height = 6, width = 6, fig.dpi=300, fig.cap=fig.diff_mort, fig.cap.sep=""}

@@

\newpage

```{r tbl-summ-grow-mort, tab.cap = tab.summ_grow_mort}

g.m.summ.ft
